Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-11 manu-draft.Rmd
shorten abstract

>>>>

From March 30th to April 9th, 2021, an estimated 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated ~205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.  An initial phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.  Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed in May, with biomass estimated at 0.5 kg/m2 at some locations.  Blooms of *Karenia brevis* were observed beginning in May and continuing through July.  Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1700 tons of dead fish.  Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region.
